---
title: "Full vs Lite Model Comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Full vs Lite Model Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(multimix)
data(sample_data)
```

## Two model variants

The `multimix` package provides two model variants that differ in their random effects structure:

### Full model (`multimix()`)

- **Two independent random effects** $u_1, u_2 \sim N(0, \sigma_1^2), N(0, \sigma_2^2)$
- $u_1$ scales the early phase, $u_2$ scales the late phase independently
- Uses 2D Gauss-Hermite quadrature for numerical integration
- **12 parameters** (before fixing): $\beta_{01}, \beta_{02}, a_1, a_2, \sigma_1, \sigma_2, t_{1/2}^{early}, t_{1/2}^{late}, \eta^{early}, \eta^{late}, \gamma^{early}, \gamma^{late}$

### Lite model (`multimix_lite()`)

- **Single shared random effect** $u \sim N(0, \sigma^2)$
- The same $u$ scales both phases (via different coefficients $a_1, a_2$)
- Uses 1D Gauss-Hermite quadrature (faster)
- **11 parameters** (before fixing): same as full model but with a single $\sigma$ instead of $\sigma_1, \sigma_2$

## When to use each

| Scenario | Recommended model |
|----------|------------------|
| Large dataset (many subjects) | Full model — can estimate two variance components |
| Small dataset (< 30 subjects) | Lite model — fewer parameters, more stable |
| Exploratory analysis | Lite model — faster convergence |
| Strong prior that phases are correlated | Lite model — shared random effect encodes this |
| Phases are driven by different mechanisms | Full model — allows independent subject variation |
| Publication/final analysis | Try both and compare via log-likelihood |

## Comparing models

```{r, eval = FALSE}
# Fit both models
full_model <- multimix(sample_data, verbose = FALSE)
lite_model <- multimix_lite(sample_data, verbose = FALSE)

# Compare log-likelihoods
cat("Full model logLik:", full_model$logLik, "\n")
cat("Lite model logLik:", lite_model$logLik, "\n")
cat("Difference:", full_model$logLik - lite_model$logLik, "\n")
```

The full model will always have a log-likelihood at least as high as the lite model (since it is a more flexible model). However, if the improvement is small, the lite model may be preferred for parsimony.

## Parameter interpretation differences

### Full model

```{r, eval = FALSE}
print(full_model)
```

In the full model, $\sigma_1$ and $\sigma_2$ represent the standard deviation of between-subject variation in the early and late phases respectively. A large $\sigma_1$ with small $\sigma_2$ suggests subjects differ mainly in their early phase response.

### Lite model

```{r, eval = FALSE}
print(lite_model)
```

In the lite model, the single $\sigma$ captures overall between-subject variation. The coefficients $a_1$ and $a_2$ determine how much of this variation is expressed in each phase.

## Visual comparison

```{r, eval = FALSE}
library(ggplot2)

p_full <- plot(full_model) + ggtitle("Full Model (two random effects)")
p_lite <- plot(lite_model) + ggtitle("Lite Model (single random effect)")

print(p_full)
print(p_lite)
```

Look for:

- **Subject-level curves**: More spread in the full model suggests the two random effects capture additional variability
- **Population average**: Should be similar between models if the data is well-behaved
- **Fit to observed data**: Compare how closely the points (empirical proportions) track the blue line

## Computational considerations

| Aspect | Full model | Lite model |
|--------|-----------|------------|
| Quadrature | 2D ($n_{GH}^2$ evaluations) | 1D ($n_{GH}$ evaluations) |
| Per-subject cost | $O(n_{GH}^2 \cdot T_i)$ | $O(n_{GH} \cdot T_i)$ |
| Typical convergence | Slower, more retries | Faster, fewer retries |
| Default nGH = 40 | 1,600 evaluations/subject | 40 evaluations/subject |

For large datasets, consider starting with the lite model to get reasonable initial parameter estimates, then using those to inform the bounds for the full model.
