---
title: "Understanding Temporal Decomposition"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding Temporal Decomposition}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(multimix)
library(ggplot2)
```

## The temporal decomposition framework

The `multimix` model decomposes the conditional odds of a binary outcome into two temporal components:

$$
\text{Conditional odds}(t) = \exp(\beta_{01} + a_1 u_1) \cdot \Lambda_1(t) + \exp(\beta_{02} + a_2 u_2) \cdot \Lambda_2(t)
$$

where $\Lambda_1(t)$ captures the **early phase** (density function $g(t)$) and $\Lambda_2(t)$ captures the **late phase** (hazard function $h(t) = g(t) / (1 - G(t))$).

Both functions are derived from a generalized temporal decomposition parameterized by:

- **Half-life** ($t_{1/2}$): controls the time scale
- **Eta** ($\eta$, called `nu` internally): exponent of time
- **Gamma** ($\gamma$, called `m` internally): shape of the distribution

## Exploring temporal basis functions

Let's visualize how the early and late phase functions behave under different parameter settings.

### Effect of half-life

```{r}
times <- seq(0.1, 20, by = 0.1)

df_thalf <- data.frame(
  Time = rep(times, 3),
  Value = c(
    get_early_phase(times, thalf = 1, eta = 1, gamma = 0),
    get_early_phase(times, thalf = 3, eta = 1, gamma = 0),
    get_early_phase(times, thalf = 8, eta = 1, gamma = 0)
  ),
  HalfLife = rep(c("t½ = 1", "t½ = 3", "t½ = 8"), each = length(times))
)

ggplot(df_thalf, aes(x = Time, y = Value, color = HalfLife)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Early Phase (Density) — Effect of Half-Life",
    x = "Time", y = "g(t)", color = "Half-Life"
  ) +
  theme_minimal(base_size = 14)
```

The half-life parameter shifts the temporal scale: shorter half-lives produce faster-decaying early phases.

### Effect of eta (time exponent)

```{r}
df_eta <- data.frame(
  Time = rep(times, 3),
  Value = c(
    get_early_phase(times, thalf = 3, eta = 0.5, gamma = 0),
    get_early_phase(times, thalf = 3, eta = 1, gamma = 0),
    get_early_phase(times, thalf = 3, eta = 3, gamma = 0)
  ),
  Eta = rep(c("η = 0.5", "η = 1", "η = 3"), each = length(times))
)

ggplot(df_eta, aes(x = Time, y = Value, color = Eta)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Early Phase (Density) — Effect of Eta",
    x = "Time", y = "g(t)", color = "Eta"
  ) +
  theme_minimal(base_size = 14)
```

Higher eta values produce sharper, more peaked density curves.

### Early vs late phase

```{r}
df_phases <- data.frame(
  Time = rep(times, 2),
  Value = c(
    get_early_phase(times, thalf = 3, eta = 1, gamma = 0),
    get_late_phase(times, thalf = 3, eta = 1, gamma = 0)
  ),
  Phase = rep(c("Early (density g(t))", "Late (hazard h(t))"), each = length(times))
)

ggplot(df_phases, aes(x = Time, y = Value, color = Phase)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Early vs Late Phase Components",
    x = "Time", y = "Value", color = "Phase"
  ) +
  theme_minimal(base_size = 14) +
  coord_cartesian(ylim = c(0, 2))
```

The early phase (density) peaks and declines, while the late phase (hazard) increases over time — reflecting the intuition that different mechanisms drive the outcome at different time scales.

## Valid parameter combinations

The `decompos()` function handles six parameter sign combinations:

| Case | $\gamma$ (m) | $\eta$ (nu) | Behavior |
|------|:---:|:---:|------|
| 1 | > 0 | > 0 | Standard sigmoidal |
| 1L | = 0 | > 0 | Limiting exponential case |
| 2 | < 0 | > 0 | Heavy-tailed |
| 2L | < 0 | = 0 | Exponential decay |
| 3 | > 0 | < 0 | Bounded cumulative |
| 3L | = 0 | < 0 | Bounded exponential |

The case where **both** $\gamma < 0$ and $\eta < 0$ is mathematically undefined and will produce an error.

## Practical guidelines

- Start with `eta = 1, gamma = 0` (exponential-like decay) for initial exploration
- Use the default bounds as a starting point and narrow them based on domain knowledge
- The half-life parameter is the most interpretable: it controls when each phase peaks or transitions
- If the optimizer struggles to converge, try fixing gamma parameters to 0 via `fixed_pars`
